|

### 2025-09-17 output/IMPS-2025-slideshow.qmd
Externalize grid transformation code
Grid transformation code was repeated in the two slideshows. To use it across the project, I externalize the code to the script with the R "Output" functions

```{r sources}
source("R/Formulae.R",                  encoding = 'UTF-8')
source("R/Output.R",                    encoding = 'UTF-8')
source("R/Mirt_toolbox.R",              encoding = 'UTF-8')
source("src/Graphical_example_paper.R", encoding = 'UTF-8')
```

@@

```{r create-grid-function}
create_grid <- function(basis,
                        x_limits = c(-4, 4),
                        y_limits = c(-4, 4),
                        break_step = 2) {
  # Argument checking and formatting: ----

  ## Basis vectors tangents and norms
  vec_tan <- basis[2, ] / basis[1, ]

  # Main: ----

  # Define grid limits as a function of the axes limits (in canonical metric):
  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
    x_lim = if_else(x == x_limits[1], "inf", "sup"),
    y_lim = if_else(y == y_limits[1], "inf", "sup")
  )

  # Select delimiting grid vertices, based on the basis vector directions
  grid_bounds <- bind_rows(
    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
    .id = "grid"
  ) |>
    filter(xor(slope > 0, x_lim == y_lim)) |>
    mutate(limit_intercept = y - slope * x) # Compute their intercept

  # Transformed (1, 1), expressed in the canonical basis
  unit_point <- basis %*% c(1, 1) |> drop() |>
    as_tibble(.name_repair = "minimal") |>
    add_column(name = c('x', 'y') |> paste0('_unit')) |>
    pivot_wider(values_from = value)

  # Grid specifications, in canonical basis
  grid_specs <- grid_bounds |>
    bind_cols(unit_point) |>
    mutate(unit_intercept = y_unit - slope * x_unit) |>
    select(-ends_with("unit")) |>
    mutate(
      n_units  = trunc(limit_intercept / unit_intercept),
      n_breaks = n_units %/% break_step * break_step
    )

  # Create vertical grid data, if the second basis vector is completely vertical
  if (is.infinite(vec_tan[2])) {

    # Create vertical grid data
    v_grid <- grid_bounds |>
      filter(grid == "v") |>
      mutate(
        unit     = basis[1, 1],
        n_units  = trunc(x / unit),
        n_breaks = n_units %/% break_step * break_step
      ) |>
      select(grid, limit, n_breaks, unit) |>
      pivot_wider(names_from = "limit", values_from = n_breaks) |>
      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
      mutate(
        grid      = 'v',
        slope     = Inf,
        intercept = NA_real_,
        label_pos = coord,
        ref       = as.character(!coord)
      )

    # Filter out non-valid grid values
    grid_specs <- grid_specs |> filter(!is.infinite(slope))
  }

  # Create grid data
  grid <- grid_specs |>
    select(grid, limit, slope, unit_intercept, n_breaks) |>
    pivot_wider(names_from = limit, values_from = n_breaks) |>
    group_by(grid) |>
    reframe(
      slope,
      coord = seq(from = inf, to = sup, by = break_step),
      intercept = coord * unit_intercept
    ) |>
    mutate(
      label_pos = if_else(
        grid == "h",
         intercept   + x_limits[1] * vec_tan[1],
        (y_limits[1] - intercept)  / vec_tan[2]
      ),
      ref = as.character(!intercept)
    )

    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))

    grid
}

plot_grid <- function(grid,
                      x_limits = c(-4, 4),
                      y_limits = c(-4, 4),
                      break_step = 2,
                      linetype_axis = 'solid',
                      linetype_grid = 'dashed') {

  # Argument checking and formatting: ----

  ## Grid plot linetypes
  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)

  ## Basis vectors tangents and norms
  vec_tan <- grid |> distinct(grid, slope) |> deframe()

  ## Axis title positions
  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
    (x_limits[2] - x_limits[1])
  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
    (y_limits[2] - y_limits[1])

  # Main: ----

  # Assign x axis label values and coordinates (to account for the vertical
  #   special case)
  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)

  # Create vertical grid geometry if the second vector is completely vertical
  if (is.infinite(vec_tan[2])) {

    geom_vgrid <- geom_vline(
      mapping = aes(xintercept = coord, linetype = ref),
      data = grid |> filter(grid == "v")
    )

    grid <- grid |> filter(grid == "h")
  }

  grid_plot <- grid |>
    ggplot() +
    geom_abline(
      mapping = aes(
        intercept = intercept,
        slope     = slope,
        linetype  = ref
      )
    )

  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid

  grid_plot +
    scale_x_continuous(
      minor_breaks = NULL,
      breaks = x_labels |> pull(label_pos),
      labels = x_labels |> pull(coord),
      limits = x_limits,
      expand = expansion(),
      name   = NULL,
      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
    ) +
    scale_y_continuous(
      minor_breaks = NULL,
      breaks = grid |> filter(grid == "h") |> pull(label_pos),
      labels = grid |> filter(grid == "h") |> pull(coord),
      limits = y_limits,
      expand = expansion(),
      name   = NULL,
      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
    ) +
    scale_linetype_manual(values = linetypes, guide = NULL) +
    coord_fixed() +
    theme(
      axis.ticks   = element_blank(),
      axis.line    = element_blank(),
      axis.title.x       = element_markdown(hjust = y_axis_titpos),
      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
    )
}

transform_grid <- function(basis,
                           x_limits = c(-4, 4),
                           y_limits = c(-4, 4),
                           break_step = 2,
                           linetype_axis = 'solid',
                           linetype_grid = 'dashed') {

  # Argument checking and formatting: ----

  # Main: ----
  basis |>
    create_grid(
      x_limits   = x_limits,
      y_limits   = y_limits,
      break_step = break_step
    ) |>
    plot_grid(
      x_limits   = x_limits,
      y_limits   = y_limits,
      break_step = break_step,
      linetype_axis = 'solid',
      linetype_grid = 'dashed'
    )
}
```

>>>>

```{r sources}
#| cache:   false
#| include: false

source("R/Formulae.R",                  encoding = 'UTF-8')
source("R/Output.R",                    encoding = 'UTF-8')
source("R/Mirt_toolbox.R",              encoding = 'UTF-8')
source("src/Graphical_example_paper.R", encoding = 'UTF-8')
```

@@
